---
title: "Project 8: Predictive Sales Analytics"
subtitle: "Objective: Use historical sales data to forecast future sales and trends"
format: pptx
editor: visual
---

```{r setup}
library(tidyverse)
library(lubridate)
library(forecast)
library(scales) # format x an y labels in ggploty
library(gt) # table formatter
library(duckdb)
library(DBI)
knitr::opts_chunk$set(echo = TRUE,message = FALSE, warning = FALSE)
```

```{r}
rm(list=ls())
```
```{r}
# Check if 'con' exists and is valid
if (exists("con") && dbIsValid(con)) {
  # Connection exists and is valid; go straight to querying
  
  sales <- dbGetQuery(con, "SELECT * FROM sales;")
  dbDisconnect(con)
} else {
  # Connection does not exist or is invalid; establish connection first
  con <- dbConnect(duckdb(), "./furniture_mart.duckdb")
  sales <- dbGetQuery(con, "SELECT * FROM sales;")
  dbDisconnect(con)
}
```

# 1. Identify seasonal trends by analyzing sales data from previous years and group them by month.

```{r}

# Ensure date column is in Date format
data <- sales %>% 
  mutate(order_date = as.Date(order_date))

# Extract year and month for seasonal analysis
data <- data %>%
  mutate(year = year(order_date),
         month = month(order_date, label = TRUE))

# Summarize monthly sales totals by year
monthly_sales <- data %>%
  group_by(year, month) %>%
  summarize(total_sales = sum(total_sales, na.rm = TRUE))
```
```{r}
# Plot with factor grouping for year without gradient scale
ggplot(monthly_sales, aes(x = month, y = total_sales, group = factor(year), color = factor(year))) +
  geom_line() +
  labs(title = "Monthly Sales Trend Over Time by Year",
       x = "Month", y = "Total Sales", color = "Year") +
  theme_minimal()

```
```{r}
# Revised faceted plot by year with ordered month and factor for year
ggplot(monthly_sales, aes(x = month, y = total_sales)) +
  geom_line(aes(color = factor(year), group = year)) +
  facet_wrap(~ factor(year), ncol = 1) +
  labs(title = "Monthly Sales Trends by Year",
       x = "Month", y = "Total Sales", color = "Year") +
  theme_minimal()
```


# 2. Generate a rolling 12-month sales forecast based on historical sales.


```{r}
library(zoo)

# Summarize and prepare monthly sales data, ensuring all months are present
monthly_sales_ts <- monthly_sales %>%
  group_by(year, month) %>%
  summarize(total_sales = sum(total_sales, na.rm = TRUE)) %>%
  ungroup() %>%
  # Combine year and month into a Date column (1st of each month)
  mutate(date = as.Date(paste(year, month, "01", sep = "-"), format = "%Y-%b-%d")) %>%
  # Ensure complete date sequence, filling missing months with NA
  complete(date = seq.Date(min(date), max(date), by = "month")) %>%
  replace_na(list(total_sales = 0)) %>%  # Fill missing sales with 0 if appropriate
  select(date, total_sales)

# Create a time series object from the cleaned data
sales_ts <- ts(monthly_sales_ts$total_sales, start = c(year(min(monthly_sales_ts$date)), month(min(monthly_sales_ts$date))), frequency = 12)

# Fit an ARIMA model to the time series
arima_model <- auto.arima(sales_ts)

# Forecast the next 12 months
forecast_sales <- forecast(arima_model, h = 12)

# Plot historical sales and forecasted sales
autoplot(forecast_sales) +
  labs(title = "12-Month Rolling Sales Forecast",
       x = "Date", y = "Total Sales") +
  theme_minimal()



```

# 3. Predict which product sub-categories are likely to see increased sales in the upcoming months

```{r}

# Convert dates to Date format if not already
online_sales <- sales %>%
  mutate(order_date = as.Date(order_date))

# Summarize monthly sales by sub_category
monthly_sales <- online_sales %>%
  group_by(sub_category, month = floor_date(order_date, "month")) %>%
  summarise(total_sales = sum(total_sales, na.rm = TRUE)) %>%
  ungroup()

# Plot monthly sales trends with trendlines for each sub_category
ggplot(monthly_sales, aes(x = month, y = total_sales, color = sub_category)) +
  geom_line(size = 1) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +  # Add trendline
  labs(title = "Monthly Sales Trends by Sub-Category with Trendlines",
       x = "Month",
       y = "Total Sales") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_wrap(~sub_category, scales = "free_y") # Separate each sub-category in a facet

# Forecast future sales for each sub_category
sub_category_forecasts <- monthly_sales %>%
  split(.$sub_category) %>%
  map(~ {
    ts_data <- ts(.x$total_sales, frequency = 12, start = c(year(min(.x$month)), month(min(.x$month))))
    fit <- auto.arima(ts_data)  # Fit an ARIMA model
    forecasted <- forecast(fit, h = 6)  # Forecast for the next 6 months
    autoplot(forecasted) + 
      ggtitle(paste("Forecast for", unique(.x$sub_category))) + 
      geom_smooth(method = "lm", se = FALSE, linetype = "dashed")  # Add trendline to forecast plot
  })

# Display forecasts
sub_category_forecasts






```

# 4. Create a forecasting report that breaks down future sales expectations by region and market  segment.

```{r}
# Load necessary libraries

# Convert dates to Date format if not already
online_sales <- sales %>%
  mutate(order_date = as.Date(order_date))

# Summarize monthly sales by region and market segment
monthly_sales_region_segment <- online_sales %>%
  group_by(region, market_segment, month = floor_date(order_date, "month")) %>%
  summarise(total_sales = sum(total_sales, na.rm = TRUE)) %>%
  ungroup()

# Initialize list to store forecast plots for each region and market segment
region_segment_forecasts <- monthly_sales_region_segment %>%
  split(., list(.$region, .$market_segment), drop = TRUE) %>%
  map(~ {
    ts_data <- ts(.x$total_sales, frequency = 12, start = c(year(min(.x$month)), month(min(.x$month))))
    fit <- auto.arima(ts_data)  # Fit an ARIMA model
    forecasted <- forecast(fit, h = 6)  # Forecast for the next 6 months
    
    # Create a dataframe of actual and forecasted values for plotting with trendline
    actual_data <- data.frame(month = .x$month, total_sales = .x$total_sales)
    forecast_data <- data.frame(month = seq(max(.x$month) + months(1), length.out = 6, by = "month"),
                                total_sales = as.numeric(forecasted$mean))
    
    combined_data <- bind_rows(actual_data, forecast_data)
    
    # Plot historical data with forecast and trendline
    ggplot(combined_data, aes(x = month, y = total_sales)) +
      geom_line(color = "blue", size = 1) +  # Historical data and forecast line
      geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red") +  # Trendline
      labs(title = paste("Forecast for Region:", unique(.x$region), 
                         "| Market Segment:", unique(.x$market_segment)),
           x = "Month",
           y = "Total Sales") +
      theme_minimal()
  })

# Display forecasts with trendlines
region_segment_forecasts



```

# 5. Suggest potential growth strategies for underperforming categories based on past data analysis.


By implementing these data-driven strategies, you can target the unique challenges and opportunities within each underperforming category, encouraging growth and expanding your market reach. Let me know if you’d like further tailored strategies based on specific sub_category performance!