---
title: "Project 8: Predictive Sales Analytics"
subtitle: "Objective: Use historical sales data to forecast future sales and trends"
format: dashboard
editor: visual
---

```{r setup}
library(tidyverse)
library(lubridate)
library(forecast)
library(scales) # format x an y labels in ggploty
library(gt) # table formatter
library(duckdb)
library(DBI)
knitr::opts_chunk$set(echo = TRUE,message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
rm(list=ls())
```

```{r,  echo=FALSE}
# Check if 'con' exists and is valid
if (exists("con") && dbIsValid(con)) {
  # Connection exists and is valid; go straight to querying
  
  sales <- dbGetQuery(con, "SELECT * FROM sales;")
  dbDisconnect(con)
} else {
  # Connection does not exist or is invalid; establish connection first
  con <- dbConnect(duckdb(), "./furniture_mart.duckdb")
  sales <- dbGetQuery(con, "SELECT * FROM sales;")
  dbDisconnect(con)
}
```

# Table of Contents

 1. Calculate the total profit and profit margin for each region.
 2. Rank regions by profitability and total sales volume.
 3. Identify regions with declining sales and provide insights on potential causes.
 4. Analyze which product categories and sub-categories are most profitable in each region.
 5. Create a report that recommends regions for increased marketing investment based on  profitability and sales growth potential.

# Page 1
#### 1. Calculate the total profit and profit margin for each region.

```{r}

# Ensure relevant columns are numeric
sales <- sales %>%
  mutate(
    profit = as.numeric(profit),
    total_sales = as.numeric(total_sales)
  )

# Calculate total profit and profit margin by region
region_summary <- sales %>%
  group_by(region) %>%
  summarise(
    total_profit = sum(profit, na.rm = TRUE),
    total_sales = sum(total_sales, na.rm = TRUE)
  ) %>%
  mutate(
    profit_margin = total_profit / total_sales
  )

# Plot: Total Profit by Region
ggplot(region_summary, aes(x = region, y = total_profit)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Total Profit by Region",
    x = "Region",
    y = "Total Profit"
  ) +
  theme_minimal()

# Plot: Profit Margin by Region
ggplot(region_summary, aes(x = region, y = profit_margin)) +
  geom_col(fill = "darkorange") +
  labs(
    title = "Profit Margin by Region",
    x = "Region",
    y = "Profit Margin"
  ) +
  theme_minimal()



```
# Page 2  
 2. Rank regions by profitability and total sales volume.
```{r}

# Ensure relevant columns are numeric
sales <- sales %>%
  mutate(
    profit = as.numeric(profit),
    total_sales = as.numeric(total_sales)
  )

# Calculate total profit, total sales, and rank regions
region_ranking <- sales %>%
  group_by(region) %>%
  summarise(
    total_profit = sum(profit, na.rm = TRUE),
    total_sales = sum(total_sales, na.rm = TRUE)
  ) %>%
  arrange(desc(total_profit)) %>%
  mutate(
    profit_rank = rank(-total_profit),
    sales_rank = rank(-total_sales)
  )

# Display the rankings
print(region_ranking)

# Plot: Rank Regions by Profitability and Sales Volume
ggplot(region_ranking, aes(x = reorder(region, -total_profit), y = total_profit)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = profit_rank), vjust = -0.5, color = "black") +
  labs(
    title = "Regions Ranked by Profitability",
    x = "Region",
    y = "Total Profit"
  ) +
  theme_minimal()

# Plot: Total Sales Volume by Region
ggplot(region_ranking, aes(x = reorder(region, -total_sales), y = total_sales)) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = sales_rank), vjust = -0.5, color = "black") +
  labs(
    title = "Regions Ranked by Total Sales Volume",
    x = "Region",
    y = "Total Sales"
  ) +
  theme_minimal()



```
 
 
# Page 3
 3. Identify regions with declining sales and provide insights on potential causes.
 
```{r}

# Ensure relevant columns are numeric and parse dates
sales <- sales %>%
  mutate(
    total_sales = as.numeric(total_sales),
    order_date = as.Date(order_date)
  ) %>%
  mutate(year = year(order_date))

# Aggregate total sales by region and year
sales_trend <- sales %>%
  group_by(region, year) %>%
  summarise(
    total_sales = sum(total_sales, na.rm = TRUE)
  ) %>%
  arrange(region, year)

# Identify declining sales trends
sales_trend <- sales_trend %>%
  group_by(region) %>%
  mutate(sales_change = total_sales - lag(total_sales)) %>%
  ungroup()

# Filter regions with declining sales
declining_regions <- sales_trend %>%
  filter(sales_change < 0)

# Plot: Sales Trends by Region
ggplot(sales_trend, aes(x = year, y = total_sales, color = region, group = region)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Sales Trends by Region",
    x = "Year",
    y = "Total Sales",
    color = "Region"
  ) +
  theme_minimal()

# Display regions with declining sales
print(declining_regions)



```

#Page 3B

**Key Features of the Analysis:**
(1) Sales Trends: Aggregates sales by region and year to detect trends over time.
(2) Declining Sales: Identifies periods where sales decreased compared to the previous year.
(3) Visualization: A line chart shows sales trends for all regions over the years.

#Page 3C

**Potential Causes:**
(1) Discounts: Check if regions with declining sales offered higher discounts.
(2) Profitability: Correlate declining sales with low or negative profit margins.
(3) Market Segments: Analyze the customer segments contributing to sales in those regions.



#Page 4
 4. Analyze which product  and sub-categories are most profitable in each region.
 
```{r}

# Data preparation
profit_summary <- sales %>%
  group_by(region, sub_category) %>%
  summarise(total_profit = sum(as.numeric(profit), na.rm = TRUE)) %>%
  arrange(region, desc(total_profit))

# Bar chart: Total profit by sub-category for each region
ggplot(profit_summary, aes(x = reorder(sub_category, total_profit), y = total_profit, fill = region)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Total Profit by Sub-Category in Each Region",
       x = "Sub-Category",
       y = "Total Profit") +
  theme_minimal()

# Scatter plot: Profit vs Sales by product, colored by sub-category for each region
sales <- sales %>% mutate(profit = as.numeric(profit), total_sales = as.numeric(total_sales))

ggplot(sales, aes(x = total_sales, y = profit, color = sub_category)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~region) +
  labs(title = "Profit vs Sales by Product in Each Region",
       x = "Total Sales",
       y = "Profit") +
  theme_minimal()


```
```{r}
# Heatmap: Profit contribution by sub-category and region
heatmap_data <- profit_summary %>%
  pivot_wider(names_from = region, values_from = total_profit, values_fill = 0)

# Convert to long format for ggplot
heatmap_data_long <- heatmap_data %>%
  pivot_longer(cols = -sub_category, names_to = "region", values_to = "total_profit")

# Create the heatmap
ggplot(heatmap_data_long, aes(x = region, y = sub_category, fill = total_profit)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Profit Contribution by Sub-Category and Region",
       x = "Region",
       y = "Sub-Category",
       fill = "Total Profit") +
  theme_minimal()
```

#Page 5

 5. Create a report that recommends regions for increased marketing investment based on  profitability and sales growth potential.

```{r}

# Ensure the dataset is loaded as 'sales' data.frame
sales <- sales %>%
  mutate(order_date = as.Date(order_date), profit = as.numeric(profit), total_sales = as.numeric(total_sales))

# Profitability Analysis by Region
profit_by_region <- sales %>%
  group_by(region) %>%
  summarise(total_profit = sum(profit, na.rm = TRUE),
            avg_profit_per_order = mean(profit, na.rm = TRUE)) %>%
  arrange(desc(total_profit))

# Sales Growth Analysis by Region (Yearly Growth)
sales_growth_by_region <- sales %>%
  mutate(year = year(order_date)) %>%
  group_by(region, year) %>%
  summarise(total_sales = sum(total_sales, na.rm = TRUE)) %>%
  arrange(region, year) %>%
  group_by(region) %>%
  mutate(yearly_growth = (total_sales / lag(total_sales) - 1) * 100) %>%
  ungroup()

# Combine Profitability and Sales Growth
combined_analysis <- profit_by_region %>%
  left_join(sales_growth_by_region %>% group_by(region) %>% summarise(avg_growth = mean(yearly_growth, na.rm = TRUE)),
            by = "region") %>%
  arrange(desc(total_profit))

# Create Recommendations
recommendations <- combined_analysis %>%
  mutate(recommendation = case_when(
    total_profit > median(total_profit, na.rm = TRUE) & avg_growth > 0 ~ "High Priority",
    total_profit > median(total_profit, na.rm = TRUE) & avg_growth <= 0 ~ "Moderate Priority",
    TRUE ~ "Low Priority"
  ))



```
```{r}
# Display the recommendations table
recommendations %>%
  select(region, total_profit, avg_profit_per_order, avg_growth, recommendation) %>%
  knitr::kable(caption = "Regional Marketing Investment Recommendations")
```